### 2013-10-27 src2/thesis.tex
update

In this thesis the focus is on data-parallelism, as opposed to task- or 
control-parallelism. Data parallelism is applicable where the input data can 
be divided into chunks and operated upon independently. \emph{more, and maybe not here}

@@

Intel ArBB is part of a movement towards thinking of structure and patterns in 
parallel programming. Other languages and libraries with similar aspirations 
are NVIDIA Thrust~\citet{THRUST} and Data.Array.Accelerate~\citet{ACCELERATEDAMP11}.
In the book ``Structured Parallel Programming'' by Michael McCool et al, a case is 
made for thinking in terms of building blocks for parallel programming~\citet{STRUCTURED}. 
They emphasise the importance of collective array operations such as map, fold, scans, 
permutations and gather/scatter operations. The book contains examples in ArBB, Cilk Plus, 
Intel Threading Building Blocks, OpenMP and OpenCL. Not all these languages include 
the patterns as abstractions, but most are possible to express in each of the languages. 
The patterns, while very powerful as abstractions, also provide tools for thinking about 
parallel algorithms.

@@

Obsidian is our embedded language for GPU programming. We raise the level of abstraction 
compared to CUDA and provide a more composable interface. But we also want to keep 
the possibility for the programmer to implement patterns in different ways. A balance 
must be struck between providing enough low-level control to get at the performance 
and still making building blocks implemented by the programmer reusable and composable. 
We want to provide more compositional tools for the programmer to experiment with while 
exploring the space of various ways to implement a given pattern or primitive. 
In chapter~\ref{chap:GPUProgramming}, there are papers describing work on Obsidian.

@@

@@

@@

@@

\emph{point to results in each paper}

\emph{provide information around the results in papers}

@@

@@

\noindent\emph{Feldspar}\citet{FELDSPAR2010} \newline
\noindent\emph{Lava}\citet{lavaICFP} \newline
\noindent\emph{Vertigo}\citet{VERTIGO}  \newline

@@

\noindent\emph{Generic monad constructs (josef + emil + anders}\citet{Generic} \newline
\noindent\emph{The Constrained Monad Problem (Gill)}\citet{sculthorpe2013constrained} \newline
\noindent\emph{Compiling Embedded Languages}\citet{COMPILEEDSL} \newline

@@

\noindent\emph{Find a niche for Obsidian}

\noindent\emph{Improve on Warp level capabilities}

\noindent\emph{Improve expressivity, code performance}

@@

\noindent\emph{What contributions are made by Obsidian} 
Looking at the related work, we se that almost all high level GPU programming 
languages provide abstractions of parallel primitives, aggregate operations, 
map-folds-zipwiths, giving justification to  our investigation of GPU programming without 
these.

\noindent\emph{What works well with Obsidian}

\noindent\emph{EmbArBB contributions}

>>>>

This thesis focuses on data-parallelism, as opposed to control- or 
task-parallelism. Data parallelism is applicable where the input data can 
be divided into chunks and operated upon independently. Data-parallelism is scalable
meaning that usually as problem size increase, the potential for parallelism increase.

@@

Intel ArBB is part of a movement towards thinking of structure and patterns in 
parallel programming. Other languages and libraries with similar aspirations 
are NVIDIA Thrust~\citet{THRUST} and Data.Array.Accelerate~\citet{ACCELERATEDAMP11}.
In the book ``Structured Parallel Programming'' by Michael McCool et al, a case is 
made for thinking in terms of building blocks for parallel programming~\citet{STRUCTURED}. 
They emphasise the importance of collective array operations such as map, fold, scans, 
permutations and gather/scatter operations. The book contains examples in ArBB, Cilk Plus, 
Intel Threading Building Blocks, OpenMP and OpenCL. Not all these languages include 
the patterns as abstractions, but most are possible to express in each of the languages. 
The patterns, while very powerful as abstractions, also provide tools for thinking about 
parallel algorithms. The ArBB library support flat regular data-parallelism, but also 
irregular and to some degree nested data-parallelism.

@@

Obsidian is our embedded language for GPU programming. We raise the level of abstraction 
compared to CUDA and provide a more composable interface. But we also want to keep 
the possibility for the programmer to implement patterns in different ways. A balance 
must be struck between providing enough low-level control to get at the performance 
and still making building blocks implemented by the programmer reusable and composable. 
We want to provide more compositional tools for the programmer to experiment with while 
exploring the space of various ways to implement a given pattern or primitive. 
In chapter~\ref{chap:GPUProgramming}, there are papers describing work on Obsidian. 
Obsidian is focused on flat and regular data-parallelism.

@@

Push arrays, like pull arrays, are functions. A Push array is a higher order 
function, taking a function from value and index to a thread program and giving 
as result a program at some level in the hierarchy. The function passed to the 
push array function we call a write-function. Below is the definition of 
push arrays that we use in Obsidian. Push arrays are parameterised on element 
type ({\tt a}), size type ({\tt s}) and program hierarchy level ({\tt p}).

@@

When working with pull arrays, the consumer of such an array decides how to 
iterate over the array and what elements to compute. This is done applying the 
push array function to those indices that are interesting to the consumer. Push 
arrays work in the opposite way. The Push array function (the result program) contain 
the iteration schema; the consumer of the push only decides what write function 
to supply. The iteration schema used is decided upon when creating a push arrays. 
The code below is an example of how to create a push array from a pull array.

\begin{verbatim} 
convertToPush :: Pull Word32 e 
               -> Push Block Word32 e 
convertToPush (Pull n ixf) =
    Push n $ \wf -> 
      forAll (fromIntegral n) $ \i -> wf (ixf i) i
\end{verbatim}

The function above takes a pull array and creates a push array with a parallel 
iteration schema over threads within a block. There are other iteration schemas to 
chose from such as sequential, combinbations of sequential and paralell and parallel 
across both threads and blocks (two level parallelism).

Push arrays give many of the same benefits that pull arrays have. Using pull 
arrays we also get fusion of operations as default. The implementation of map 
(shown below) illustrates how maping a function {\tt f} over a push array results 
in an application of {\tt f} in the write-function.

\begin{verbatim} 
map f (Push s p) = Push s $ \wf -> p (\e ix -> wf (f e) ix)
\end{verbatim}

@@

When compiling Obsidian programs to CUDA, we only need to be concerned with the 
{\tt Program} datatype. Given the hierarchy level type parameters to the program datatype 
we know that the AST can be quite directly translated to CUDA. We know that there will 
be at most one level of nestedness in the parallelism (several blocks, each of several threads). 
This means that we do not need to do any transformations on the AST before generating CUDA code.

@@

We have implemented basic algorithms such as sorting, merging, fractal generation, reductions and parallel prefix (scans). Often we have managed to get very good performance, close to that 
of hand optimised CUDA code. However, there are also cases where we just cant seem to reach 
the level of performance that we would want. One such example is parallel prefix, an 
application that has been haunting us from the very beginning. Scan is problematic to us because 
it is best implemented as an in-place algorithm while our array representations do 
not allow for in-place updated (there is no memory associated with them to update). 
We are trying to improve this situation by adding a third array representation to Obsidian,
mutable arrays.

Paper F (section~\ref{sec:paperF}), tells a story about how to implement and tune 
a reduction kernel in Obsidian. At the end the implemented reduction kernel is 
faster than the reduction skeleton used in Accelerate. This is quite a feat, since 
reduction in accelerate corresponds to just one single hand optimised skeleton. We would have 
been happy just getting close.

In paper E (section~\ref{sec:paperE}), we implement counting sort and occurrance sort 
using Obsidian. We use very low-level features of Obsidian and end up producing 
code that performs very well.

Paper D (section~\ref{sec:paperD}), introduces push arrays and uses them in the 
implementation of sorting networks to quite good results. With push arrays we 
where able to experiment with more detailed decompositions (what thread computes 
what value) of the algorithms.

@@

EmbArBB provides many of the ArBB operations on Arrays butprovides a Repa like 
shape-polymorphic interface to them. For example a reduction operation can take a two  
dimensional vector and reduce all rows or columns and provide a one dimensional result. Or 
it takes a one dimensional vector and returns a single value (zero dimensional vector).

\begin{verbatim} 
addReduce :: Num a 
           => Exp USize 
           -> Exp (DVector (t:.Int) a) 
           -> Exp (DVector t a) 
\end{verbatim}

The {\tt addReduce} function takes an input vector that is atleast one dimensional 
({\tt t:.Int}) and returns a vector that has one less dimension ({\tt t}).

@@

Feldspar is an embedded language for digital signal processors~\citet{FELDSPAR}. Apart 
from using the syntactic library~\citet{SyntacticICFP12}, it is quite similar to Obsidian. 
There is a deeply embedded core language (called Core) and vectors as a shallow embedding 
on top of that, just like in Obsidian.

When we first started working on Obsidian we used Lava~\citet{LavaICFP} as a inspiration. 
We wanted to mimic the Lava programming style but generate GPU code. At first this worked quite 
well and most Obsidian programs superficially looked quite like Lava programs. We have diverged 
from this aspiration since realising that GPUs are quite a different story and performance 
requires more control of GPU architecture specifics.

@@

In paper A~(\ref{sec:paperA}) we describe our method for compiling monadic embedded languages. 
This means we need to take a monadic computation and observe its structure in order to generate
code. This work is closely related to a method described in ``The Constrained Monad Problem''
by Sculthorpe and Gill~\citet{sculthorpe2013constrained}. In this paper they solve a more 
general problem than just compilation but our method is nice because it is easy to understand 
and solves a single problem and does it well. In ``Generic monad constructs'' Persson et al 
also solves the problem of compiling monads using a continuation monad.

``Compiling Embedded Languages'' by Elliott and Finne has been a constant source of information 
and something we have refered to in almost all our work. This is, in my opinion, the paper to turn 
to for anyone starting out with compiled embedded languages.

@@

Obsidian is a language for GPU kernel implementation. It is not as well integrated 
into Haskell as for example Accelerate. There are ways to execute Obsidian kernels 
from within Haskell; the recommended workflow, however, is to generate the CUDA  
as text and write the host code by hand. This can be seen both a shortcoming 
and a feature. It is hard to integrate Accelerate programs in larger applications 
written in other languages. While when using Obsidian a kernel is generated and 
can be used as part of for example a library.

Obsidian is also a quite low-level language. The programmer needs to know GPU details 
and is allowed to write programs at a level almost as low as CUDA, if it is desired. 
Paper E (section~\ref{sec:paperE}) contains examples of programming at this low level. 
One potential use of Obsidian could be as the code generating backend of high-level 
and very domain specific embedded languages. This could be a niche for Obsidian.

Here and there in the thesis it has been hinted that Obsidian has warp level 
capabilites and that these are not as integrated into the hierarchy as block, 
thread and grid are. The warp abstractions offered also requires more work in 
the code generating backend than the others, that correspond directly to CUDA concepts. 
The problem is that CUDA does not have a warp abstraction. Warps needs to be 
better integrated with the rest of Obsidian and the code generation required 
for warps needs more testing before being trustworthy and useful.

%\noindent\emph{Improve on Warp level capabilities}

%\noindent\emph{Improve expressivity, code performance}

@@

This thesis contains two approaches to embedding a language for data-parallel programming 
in Haskell. EmbArBB embeds a preexisting library for data-parallelism in Haskell 
and at little effort from us, very good performance is gained. The performance of 
EmbArBB is entirely due to all the optimisation expertice and effort that invested in 
the ArBB system. We get all that expertice for free. What EmbArBB provide is a 
way for Haskellers to benefit from the ArBB system. ArBB and EmbArBB is based on a 
set of parallel primitives for the programmer to use and compose. Obsidian takes 
a different approach. Built in primitives are very few and very low level, such as 
sequential and parallel for loops. On top of the low-level capabilities a library 
of high-level primitives can be built. The key is that we want the programmer to 
be able to specify how to implement these higher level abstractions.

Looking at the related work, we see that almost all high level GPU programming 
languages provide abstractions of parallel primitives, aggregate operations, 
maps, folds and reductions, giving justification to our investigation of GPU 
programming without these. We believe that the GPU architecture is still volatile 
enough that there be no ``the right'' way of implementing these primitives. As of 
today, the programmer still needs to explore the design space and find a sollution 
that performs well on a particular GPU.

Performance of Obsidian kernels is often good and the Obsidian code that they 
are generated from is often shorted than corresponding CUDA programs. Obsidian 
programs differ from CUDA programs by describing whole programs. With this I mean 
that a CUDA programmer needs to take a leap from the single threaded program 
parameterised on thread Id, to what actually will happen when this program is 
run by hundreds or thousands threads at once. The extensive indexing arithmetic 
often present in CUDA programs can make this leap a hard one. Similar programs 
in Obsidian could use operations on whole arrays such as {\tt map} or explicitly 
bounded parallel loops.

Push arrays where invented by Koen Claessen and first implemented in Obsidian 
to provide more control over where computation takes place (which thread computes 
which element). Push arrays has since been experimented on by others and show up 
in Nikola~\citet{NIKOLAPUSH}, been experimented upon by the people behind Feldspar 
and appear in StreamHS~\citet{FPCDSL}.
